Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests

### 2025-05-01 manuscript.tex
Mamuns changes

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
\title{Detecting Interactions in Random Forest for Genomic Data}

\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
\ead{halmamun@intergrain.com}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
\affiliation{organization={},
             addressline={},
             city={},
             postcode={},
             state={},
             country={}}
%% \fntext[label3]{}

@@

\author{} %% Author name

%% Author affiliation
\affiliation{organization={},%Department and Organization
            addressline={}, 
            city={},
            postcode={}, 
            state={},
            country={}}

@@

\author{Klara Verbyla}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={306 Carmody Rd}, 
  city={St Lucia},
  postcode={4067}, 
  state={QLD},
  country={Australia}}

@@

%% Keywords
\begin{keyword}
Epistatic interactions \sep  Random Forest \sep SNP data

@@

% \abstract{Detecting epistatic interactions is very important because it will help us to understand the genetic mechanism
%   on phenotypic variation. Finding epistatic interactions from high density SNP data is very challenging due to the
%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forest models were
%   widely used in a variety of disciplines including epistatic interactions. Although Random Forest models account for
%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forest based
%   methods to detect interactions relay on some kind of variable importance measures that suffer when the interacting
%   variables have very small or no marginal effects. We proposed a Random Forest based method that is not based on
%   variable importance and can find epistatic interactions from high density SNP markers in reasonable time. The proposed
%   method was first applied on several simulation datasets representing different data structures with respect to SNP
%   markers and heritibilities. Once the method was validated for finding epistatic interactions from simulated SNP data,
%   it was applied on two different traits (case-control data rheumatoid arthritis, and quantitative traits of mouse body
%   weight) in human and mouse, respectively. In both cases, the method identified several genes closely positioned to the
%   interacting SNPs that potentially have effects on the respective traits.}  \keywords{Epistatic interactions, Random
%   Forest, SNP data}

@@

% Random Forests are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble learning
% combines multiple learning algorithms/models to improve prediction or classification performance beyond what could be
% obtained from the constituent models.  In a Random Forest, \textit{m} random samples (with replacement) are selected by
% bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break
% down the correlation between features, only a random subset of features is considered at any node of the tree building
% process. For classification problems, the class of the validation data is predicted based on majority voting and for the
% regression problem, the model outputs the mean of the individual trees prediction. There are two main parameters in the
% implementation of Random Forest algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
% randomly selected features (\texttt{mtry}) from which the best split is selected at any given node.

@@

Random Forest models are widely used \cite[]{lundbergConsistentIndividualizedFeature2019} in a variety of
disciplines. The attraction of the model is based on several attributes, namely:
\begin{itemize} 
\item a high degree of accuracy;
\item a resistance to overfitting, including in the presence of large numbers of redundant variables; 
\item the provision of an out-of-bag error measure; 
\item the provision of a variable ranking at little extra computational cost.
\end{itemize}

In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forest based
approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forest-based approaches for identifying
interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
interaction or simply two variables with strong marginal effects. Additionally, if the interacting variables have small
marginal effects, they will not appear near the top of the variable importance list. This means that detecting
interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
tests.  \cite{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
other QTL mapping methods at identifying QTL and discrete trait associations. They provided benchmarks on simulated and
real yeast cross data indicating that the RF-based methods outperform other commonly used approaches for detecting
epistasis.

To overcome the challenge of a large number of pairwise SNP interaction tests, we proposed an effective two-step process
based on our observations from using a Random Forest approach. Using simulation data, we observed that ``if two SNPs are
interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
not interacting''.

% This observation was used to narrow down the vast search space of pairwise tests to detect SNP
% interactions. First, we built a Random Forest and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we selected only the SNP pairs that occurred more times that would
% be expected by chance and exceeded a set threshold of \textit{t}. In our study, we selected all the SNP pairs those have
% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forest.  Finally, using
% a linear model, each selected SNP pair was tested to determine whether the interaction was significant.

@@

The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
available genomic data. For both the simulated and real data examples, the R package ranger
\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
%Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
%the decision trees were recorded,
Working from the random forest output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices and set the diagonal to 0,
we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.  This is likely to be large and contain a large number of
zeros ($4 \times 10^9$ entries and 0.999 zero in the rheumatoid example, section \ref{subsection: rheumatoid}).

@@

Using the Rheumatoid data as an example, we have $p= 94485$ and after fitting a random forest with
1000 trees we see about 290 non-terminal splits per tree. We observe 291142 splits with a distribution of frequencies
given in the ``Observed'' column of Table \ref{tab:expected_vs_observed}.

@@

Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forest with more
trees and we will observe more counts.
There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
each variable.

@@

This, despite some processing of the linear models, is less time consuming than training a much larger Random Forest.

@@

% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forest model was created using all avaiable data. Then for each pair 
% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
% \#SNPs/3.
% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
% on the values of another independent variable. So our variables in the model can be divided between main effects and
% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
% product of the two genotype values, using the following linear regression model:

@@

\subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
\label{subsection: rheumatoid}
The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)
\cite[]{burtonGenomewideAssociationStudy2007brief}. The original dataset contains 14,000 cases for seven complex human
diseases and 3,000 shared controls. The control individuals were from two different sources: 1958 British Birth Cohort
(58C) and the UK Blood Service Control Group (UKBS). Each of the control group has 1,500 individuals. Samples of DNA
from all 17,000 individuals were genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip).  The current
analysis only used the data for rheumatoid arthritis (RA) that consisted of 4,798 individuals with 1,860 cases and 2,938
controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forest
(RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
\cite{mengPerformanceRandomForest2009}), the dataset was also filtered based on LD. Using PLINK 1.9
\cite[]{purcellPLINKToolSet2007} and a sliding window of 200 kb, one SNP from a pair of SNPs in that window was removed
if pairwise $r^2 >=0.3$.

@@

% \subsection{The Proposed Method -- count co-occurrences}
% To overcome the challenge of a large number of pairwise interaction tests, we proposed an effective two-step process
% based on our observations from using Random Forest. Using different simulation data, we observed that ``if two SNPs are
% interacting, they appear as parent-child nodes in a decision tree of a random forest more frequently than if they are
% not interacting''.  So, we utilised this observation to narrow down the vast search space of pairwise tests to detect
% interactions. First, we build a random forest and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we select only the SNP pairs that occur more times that would be
% expected by chance and exceed the set threshold. Finally, using a linear model, each selected SNP pair is tested to see
% if the interaction is significant.

\section{Results and Discussion}
\subsection{simulated data}
% Based on publicly available pig genotype data, simulated phenotypes using different scenarios with various combinations
% of number of SNPs, number of interacting SNP pairs, heritability, QTL effect size and interaction effect size were used.
% The performance of the method was assessed based on the proportion of time the method selected the interacting SNP
% pairs for pairwise interaction tests. Table \ref{table:ranking10qtl} shows the importance ranking of three interacting
% SNP pairs based on the number of times they appear as parent-child in the Random Forest decision trees. Higher ranking,
% represented by a smaller number, means a higher count. For example, a value of 1 for pair P3 means that SNP pair 3
% appeared as parent-child more than any other SNP pair in that Random Forest. Since in the base model three pairs of
% interacting SNPs do not have any marginal effects or any interaction effects, they did not appear as parent-child in any
% decision tree of the Random Forest. In the no interaction model, the three interacting SNP pairs appeared as
% parent-child in some of the scenarios. This is because in the no interaction model, although the three pairs do not
% have any interaction effect, they have their own marginal effects. Hence, they were selected at different nodes of the
% decision trees. For the interaction model, rankings of the three interacting SNP pairs are relatively higher than the
% other two models. This ranking table demonstrates that if we rank the SNP pairs based on the count of their appearance
% as parent-child, we only need to test pairwise interaction for a small amount of SNP pairs from the vast number of total
% possible combinations of SNP.

@@

% In the proposed two step method for detecting SNP-SNP interactions, a random forest model was built and the count
% determined for how many times a given pair of SNPs appears as parent-child. In the second step, only those SNP pairs
% having the counts higher than a certain number t were tested. In this simulation, we set t = 2, meaning that if any of
% the SNP pairs appeared as parent-child in the decision trees of the Random Forest for at least 2 times, we selected them
% to test for pairwise interaction using Equation 1. The third column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of pair-wise interaction tests performed
% with the interaction model for different combinations of heritability and number of SNPs. The fourth column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of possible pairwise tests for a given
% number of SNPs. From the last column of the Table \ref{table:Number.of.tests.performed.to.detect.interactions} we can
% see that, using our method we were able to narrow down the search space of pairwise test significantly. For all
% scenarios, less than one percentage of SNP pairs were tested for interaction.

@@

% \subsection{Effect of heritability and interactions p-value on the proposed method to detect candidate SNP pairs}
% Figure \ref{figure:Eff_Trait_h2_simulated_data.png} shows the pairwise counts for each of the interacting SNP pairs in
% each Random Forest at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
% of interacting SNPs. At low heritability, the pairwise count is smaller and as heritability increases, the pairwise
% count also increases. Although the three interacting SNP pairs have very similar interaction effects (ranged between 0.3
% to 0.4), they have very different numbers of counts. There is a direct relationship between the $p$-value of the
% interaction term with the pairwise count. Lower $p$-values of the interaction terms have higher pairwise count
% (correlation coefficient = 0.84). This is particularly interesting as even very small interaction effects with high
% significance will not be overlooked because of higher pairwise count.

\subsection{heritability}

@@

\subsubsection{Rheumatoid arthritis data}
% After pre-processing and filtering, the RA dataset contained 94,486 SNPs (~4.46 billion possible pairwise interaction
% tests) and 4,798 individuals.  Using random forest, we have selected only 654 SNP pairs (out of more than 4.46 billion
% possible pairs) for pairwise interaction tests (Equation 2). Among the 654 SNP pairs, 26 pairs have significant
% Bonferroni corrected interaction $p$-values ($p < 0.05$). In detail, 24 SNPs were involved in 26 significant interactions
% among which 22 SNPs were from chromosome 6 and two other from chromosome 1 and chromosome 7. Furthermore, several SNPs
% from chromosome 6 were involved in multiple interactions. All the 22 SNPs in chromosome 6 are in the HLA (human
% leukocyte antigen) region spanning from 31.35 MB to 32.80 MB. The HLA region in human genome is known to contain very
% high linkage disequilibrium (LD). If interacting SNP pairs are in the same haplotype or LD block, the significant
% interaction between the pairs might be a haplotype effect on RA, rather than true interaction. Out of 26 pairs of
% interacting SNPs, 12 pairs (all are in chromosome 6) have moderate to high LD (D’ = (0.26 to 0.55)) between the
% pairs. Hence, we retained only 14 SNP pairs which has low LD (D’ $\leq$ 0.2).

@@

\section{Conclusion}
% In this study, we proposed a simple and easy method to select candidate SNP pairs to test for interactions in large
% genomic data. We used a Random Forest algorithm to select a subset of SNP pairs from a vast search space so that the
% selected SNP pairs can be tested for interaction in a reasonable computational time. The method was validated with
% several simulation datasets representing different data structures (with respect to SNP marker density and
% heritabilities). We then used the method to identify SNP interactions for two different traits (case-control data
% rheumatoid arthritis, and quantitative traits of mouse body weight) in human and mouse, respectively. In both cases, the
% method identified several genes closely positioned to the interacting SNPs that potentially have effects on the
% respective traits. Although the approach requires further biological validation, candidate SNP pair markers identified
% by this method will provide valuable information and a genetic basis for understanding a complex trait that is the
% result of effects from multiple genes and their complex interactions. We only used SNP data to demonstrate our proposed
% method. However, the method is equally applicable when the datasets contain both SNP data and environmental data.

>>>>

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
\title{Detection of Epistatic Interactions in High-Dimensional Genomic Data Using Random Forests}

\author{Hawlader A. Al-Mamun\corref{cor1}\fnref{label2}}
\ead{hawlader.almamun@uwa.edu.au}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={2 - 40 Clunies Ross Street}, 
  city={Acton},
  postcode={2601}, 
  state={ACT},
  country={Australia}}

\affiliation{organization={InterGrain Pty Ltd.},%Department and Organization
  addressline={19 Ambitious Link}, 
  city={Bibra Lake},
  postcode={6163}, 
  state={WA},
  country={Australia}}

\affiliation{organization={School of Biological Sciences},%Department and Organization
  addressline={The University of Western Australia}, 
  city={Crawley},
  postcode={6009}, 
  state={WA},
  country={Australia}}

%% \fntext[label3]{}

@@

@@

\author{Klara Verbyla}
\affiliation{organization={Agriculture \& Food, CSIRO},%Department and Organization
  addressline={2 - 40 Clunies Ross Street}, 
  city={Acton},
  postcode={2601}, 
  state={ACT},
  country={Australia}}

@@

%% Keywords
\begin{keyword}
Epistatic interactions \sep  Random Forests \sep SNP data

@@

% \abstract{Detecting epistatic interactions is very important because it will help us to understand the genetic mechanism
%   on phenotypic variation. Finding epistatic interactions from high density SNP data is very challenging due to the
%   computational burden induced by the combinatorial number of tests that have to be performed. Random Forests models were
%   widely used in a variety of disciplines including epistatic interactions. Although Random Forests models account for
%   variable interactions, it does not explicitly detect variable interactions. And most of the Random Forests based
%   methods to detect interactions relay on some kind of variable importance measures that suffer when the interacting
%   variables have very small or no marginal effects. We proposed a Random Forests based method that is not based on
%   variable importance and can find epistatic interactions from high density SNP markers in reasonable time. The proposed
%   method was first applied on several simulation datasets representing different data structures with respect to SNP
%   markers and heritibilities. Once the method was validated for finding epistatic interactions from simulated SNP data,
%   it was applied on two different traits (case-control data rheumatoid arthritis, and quantitative traits of mouse body
%   weight) in human and mouse, respectively. In both cases, the method identified several genes closely positioned to the
%   interacting SNPs that potentially have effects on the respective traits.}  \keywords{Epistatic interactions, Random
%   Forest, SNP data}

@@

% Random Forests are an ensemble learning method first proposed by Leo Breiman \cite[]{Breiman2001}. Ensemble learning
% combines multiple learning algorithms/models to improve prediction or classification performance beyond what could be
% obtained from the constituent models.  In a Random Forests, \textit{m} random samples (with replacement) are selected by
% bootstrap sampling. Then multiple decision trees are created based on the bootstrap samples. To break
% down the correlation between features, only a random subset of features is considered at any node of the tree building
% process. For classification problems, the class of the validation data is predicted based on majority voting and for the
% regression problem, the model outputs the mean of the individual trees prediction. There are two main parameters in the
% implementation of Random Forests algorithm namely the number of decision trees (\texttt{num.trees}) and the number of
% randomly selected features (\texttt{mtry}) from which the best split is selected at any given node.

@@

Random Forests models are widely used \cite[]{lundbergConsistentIndividualizedFeature2019} in a variety of
disciplines. The attraction of the model is based on several attributes, namely:
\begin{itemize} 
\item a high degree of accuracy;
\item a resistance to overfitting, including in the presence of large numbers of redundant variables; 
\item the provision of an out-of-bag error measure; 
\item the provision of a variable ranking at little extra computational cost.
\end{itemize}

In addition it is clear that Random Forests can perform well with correlated and interacting variables, at least in the
sense that the fitted model performs well \cite[]{Ziegler.et.al.2007}. However, most of the Random Forests based
approaches are either based on single variable importance \cite[]{Jiang.et.al.2009, yoshidaSNPInterForestNewMethod2011}
or based on pairwise variable importance \cite[]{Ziegler.et.al.2007}. The Random Forests-based approaches for identifying
interactions based on variable importance measurements cannot distinguish whether a discovered interaction is a true
interaction or simply two variables with strong marginal effects. Additionally, if the interacting variables have small
marginal effects, they will not appear near the top of the variable importance list. This means that detecting
interactions based on variable importance can be problematic. It is also heavily dependent on extensive permutation
tests.  \cite{schmalohrDetectionEpistaticInteractions2018} notes that there is a lack of accurate methods for the
detection of interactions between genomic loci that affect QTLs but that several studies have shown that RF outperforms
other QTL mapping methods at identifying QTL and discrete trait associations. They provided benchmarks on simulated and
real yeast cross data indicating that the RF-based methods outperform other commonly used approaches for detecting
epistasis.

To overcome the challenge of a large number of pairwise SNP interaction tests, we proposed an effective two-step process
based on our observations from using a Random Forests approach. Using simulation data, we observed that \textit{``if two SNPs are
interacting, they appear as parent-child nodes in a decision tree of a Random Forests more frequently than if they are
not interacting''}.

% This observation was used to narrow down the vast search space of pairwise tests to detect SNP
% interactions. First, we built a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we selected only the SNP pairs that occurred more times that would
% be expected by chance and exceeded a set threshold of \textit{t}. In our study, we selected all the SNP pairs those have
% at least two (\textit{t}=2) appearance as parent-child nodes in the decision trees of the Random Forests.  Finally, using
% a linear model, each selected SNP pair was tested to determine whether the interaction was significant.

@@

The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model is created using all
available genomic data. For both the simulated and real data examples, the R package ranger
\cite[]{Wright.and.Ziegler.2017} was used with the parameters \texttt{num.tree} = 1000 and \texttt{mtry} = \#SNPs/3.
%Once the Random Forests model is built, the number of times each pair of markers was found as parent-child nodes in all
%the decision trees were recorded,
Working from the Random Forests output, we can form a $p \times p$ matrix of values such that the $\{i,j\}^{th}$ entry is the count of the
number of times variable $v_i$ is followed by variable $v_j$.  If we sum the upper and lower triangular matrices and set the diagonal to 0,
we form the interaction vector, of length $\frac {p(p-1)} 2 = {p \choose 2}$.  This is likely to be large and contain a large number of
zeros ($4 \times 10^9$ entries and 0.999 zero in the rheumatoid example, section \ref{subsection: rheumatoid}).

@@

Using the Rheumatoid data as an example, we have $p= 94485$ and after fitting a Random Forests with
1000 trees we see about 290 non-terminal splits per tree. We observe 291142 splits with a distribution of frequencies
given in the ``Observed'' column of Table \ref{tab:expected_vs_observed}.

@@

Now, somewhat uniquely, we can move the contaminating distribution. All we need to do is fit a Random Forests with more
trees and we will observe more counts.
There is a trade-off here and as we are sampling a space with typically more than $10^{9}$ possible interactions,
we have to increase the number of trees a lot to get a meaningful difference in the mean of the contaminating
distribution. Because of this we resort to a threshholding stratey that is both computationally easier and controls the FDR rate by testing
each variable.

@@

This, despite some processing of the linear models, is less time consuming than training a much larger Random Forests.

@@

% The proposed method is described in Figure \ref{figure:RF_Method}. First, a Random Forests model was created using all avaiable data. Then for each pair 
% of SNP, how many times they appeared as parent-child node in each of the decision trees were recorded. Then the SNP pairs were ranked based on
% the count and only the pairs having count >= 2 were retained. Finally, using the linier equation the interaction term was tested. In our
% studies, we used the ranger R package \cite[]{Wright.and.Ziegler.2017} and we selected \texttt{num.tree} = 1000 and \texttt{mtry} =
% \#SNPs/3.
% For a linear model, an interaction occurs when an independent variable has a different effect on the outcome depending
% on the values of another independent variable. So our variables in the model can be divided between main effects and
% interaction effects. A multiplicative epistasis model was used, where the interaction between two SNPs, $i$ and $j$,  were modelled as the
% product of the two genotype values, using the following linear regression model:

@@

\subsection{Real datasets -- Wellcome Trust Case Control Consortium data for rheumatoid arthritis}
\label{subsection: rheumatoid}
The first dataset was obtained from the Wellcome Trust Case Control Consortium (WTCCC)
\cite[]{burtonGenomewideAssociationStudy2007brief}. The original dataset contains 14,000 cases for seven complex human
diseases and 3,000 shared controls. The control individuals were from two different sources: 1958 British Birth Cohort
(58C) and the UK Blood Service Control Group (UKBS). Each of the control group has 1,500 individuals. Samples of DNA
from all 17,000 individuals were genotyped with the GeneChip 500K Mapping Array Set (Affymetrix chip).  The current
analysis only used the data for rheumatoid arthritis (RA) that consisted of 4,798 individuals with 1,860 cases and 2,938
controls. Variants having $>10\%$ missing values and MAFs$<0.05$ were removed. SNP markers that failed the
Hardy-Weinberg test (p $\leq$ 0.05) were also removed. As linkage disequilibrium (LD) could affect how the Random Forests
(RF) algorithm selects a SNP at a particular node (for more detail how LD can affect RF, please see
\cite{mengPerformanceRandomForest2009}), the dataset was also filtered based on LD. Using PLINK 1.9
\cite[]{purcellPLINKToolSet2007} and a sliding window of 200 kb, one SNP from a pair of SNPs in that window was removed
if pairwise $r^2 >=0.3$.

@@

% \subsection{The Proposed Method -- count co-occurrences}
% To overcome the challenge of a large number of pairwise interaction tests, we proposed an effective two-step process
% based on our observations from using Random Forests. Using different simulation data, we observed that ``if two SNPs are
% interacting, they appear as parent-child nodes in a decision tree of a Random Forests more frequently than if they are
% not interacting''.  So, we utilised this observation to narrow down the vast search space of pairwise tests to detect
% interactions. First, we build a Random Forests and for each pair of SNPs we count the total number of times any SNP pair
% occurred as parent-child nodes in this forest. Then we select only the SNP pairs that occur more times that would be
% expected by chance and exceed the set threshold. Finally, using a linear model, each selected SNP pair is tested to see
% if the interaction is significant.

\section{Results and Discussion}
\subsection{Simulated data}
% Based on publicly available pig genotype data, simulated phenotypes using different scenarios with various combinations
% of number of SNPs, number of interacting SNP pairs, heritability, QTL effect size and interaction effect size were used.
% The performance of the method was assessed based on the proportion of time the method selected the interacting SNP
% pairs for pairwise interaction tests. Table \ref{table:ranking10qtl} shows the importance ranking of three interacting
% SNP pairs based on the number of times they appear as parent-child in the Random Forests decision trees. Higher ranking,
% represented by a smaller number, means a higher count. For example, a value of 1 for pair P3 means that SNP pair 3
% appeared as parent-child more than any other SNP pair in that Random Forests. Since in the base model three pairs of
% interacting SNPs do not have any marginal effects or any interaction effects, they did not appear as parent-child in any
% decision tree of the Random Forests. In the no interaction model, the three interacting SNP pairs appeared as
% parent-child in some of the scenarios. This is because in the no interaction model, although the three pairs do not
% have any interaction effect, they have their own marginal effects. Hence, they were selected at different nodes of the
% decision trees. For the interaction model, rankings of the three interacting SNP pairs are relatively higher than the
% other two models. This ranking table demonstrates that if we rank the SNP pairs based on the count of their appearance
% as parent-child, we only need to test pairwise interaction for a small amount of SNP pairs from the vast number of total
% possible combinations of SNP.

@@

% In the proposed two step method for detecting SNP-SNP interactions, a random forests model was built and the count
% determined for how many times a given pair of SNPs appears as parent-child. In the second step, only those SNP pairs
% having the counts higher than a certain number t were tested. In this simulation, we set t = 2, meaning that if any of
% the SNP pairs appeared as parent-child in the decision trees of the Random Forests for at least 2 times, we selected them
% to test for pairwise interaction using Equation 1. The third column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of pair-wise interaction tests performed
% with the interaction model for different combinations of heritability and number of SNPs. The fourth column of Table
% \ref{table:Number.of.tests.performed.to.detect.interactions} shows the number of possible pairwise tests for a given
% number of SNPs. From the last column of the Table \ref{table:Number.of.tests.performed.to.detect.interactions} we can
% see that, using our method we were able to narrow down the search space of pairwise test significantly. For all
% scenarios, less than one percentage of SNP pairs were tested for interaction.

@@

% \subsection{Effect of heritability and interactions p-value on the proposed method to detect candidate SNP pairs}
% Figure \ref{figure:Eff_Trait_h2_simulated_data.png} shows the pairwise counts for each of the interacting SNP pairs in
% each Random Forests at different heritabilities (h = 0.33, 0.50, 0.66, and 1.00) for 10,000 SNP with 10 QTL and 3 pairs
% of interacting SNPs. At low heritability, the pairwise count is smaller and as heritability increases, the pairwise
% count also increases. Although the three interacting SNP pairs have very similar interaction effects (ranged between 0.3
% to 0.4), they have very different numbers of counts. There is a direct relationship between the $p$-value of the
% interaction term with the pairwise count. Lower $p$-values of the interaction terms have higher pairwise count
% (correlation coefficient = 0.84). This is particularly interesting as even very small interaction effects with high
% significance will not be overlooked because of higher pairwise count.

\subsection{Heritability}

@@

\subsubsection{Rheumatoid arthritis data}
% After pre-processing and filtering, the RA dataset contained 94,486 SNPs (~4.46 billion possible pairwise interaction
% tests) and 4,798 individuals.  Using Random Forests, we have selected only 654 SNP pairs (out of more than 4.46 billion
% possible pairs) for pairwise interaction tests (Equation 2). Among the 654 SNP pairs, 26 pairs have significant
% Bonferroni corrected interaction $p$-values ($p < 0.05$). In detail, 24 SNPs were involved in 26 significant interactions
% among which 22 SNPs were from chromosome 6 and two other from chromosome 1 and chromosome 7. Furthermore, several SNPs
% from chromosome 6 were involved in multiple interactions. All the 22 SNPs in chromosome 6 are in the HLA (human
% leukocyte antigen) region spanning from 31.35 MB to 32.80 MB. The HLA region in human genome is known to contain very
% high linkage disequilibrium (LD). If interacting SNP pairs are in the same haplotype or LD block, the significant
% interaction between the pairs might be a haplotype effect on RA, rather than true interaction. Out of 26 pairs of
% interacting SNPs, 12 pairs (all are in chromosome 6) have moderate to high LD (D’ = (0.26 to 0.55)) between the
% pairs. Hence, we retained only 14 SNP pairs which has low LD (D’ $\leq$ 0.2).

@@

\section{Conclusion}
% In this study, we proposed a simple and easy method to select candidate SNP pairs to test for interactions in large
% genomic data. We used a Random Forests algorithm to select a subset of SNP pairs from a vast search space so that the
% selected SNP pairs can be tested for interaction in a reasonable computational time. The method was validated with
% several simulation datasets representing different data structures (with respect to SNP marker density and
% heritabilities). We then used the method to identify SNP interactions for two different traits (case-control data
% rheumatoid arthritis, and quantitative traits of mouse body weight) in human and mouse, respectively. In both cases, the
% method identified several genes closely positioned to the interacting SNPs that potentially have effects on the
% respective traits. Although the approach requires further biological validation, candidate SNP pair markers identified
% by this method will provide valuable information and a genetic basis for understanding a complex trait that is the
% result of effects from multiple genes and their complex interactions. We only used SNP data to demonstrate our proposed
% method. However, the method is equally applicable when the datasets contain both SNP data and environmental data.
